---
title: "Engsci 211 Task 1: Sleep Deprivation"
author: "Alex Verkerk"
output:
  html_document:
    df_print: paged
  word_document: default
---
```{r}
library(s20x)
#laptop:/
#setwd("C:/Users/Alex/Documents/GitHub/Engsci211Assignment2")
#home computer:
setwd("I:/GitHub/Engsci 211/Engsci211Assignment2")
sleepDep.df = read.table("sleepDeprivation.txt", header = TRUE)
head(sleepDep.df)
#Deprived is already a factor with the states yes and no
```

### **Exploratory analysis**  
```{r}
boxplot(SleepHrs ~ Deprived, data = sleepDep.df)
summaryStats(SleepHrs ~ Deprived, data = sleepDep.df)
```
* Centre: Mean sleep hours for deprived is lower than mean sleep hours for non deprived
* Spread: The spread is slightly wider for non deprived as compared to deprived
* Skew: Both Yes and No appear fairly symmetrical without too significant of a tail, so no significant skew
```{r}
layout20x(1,2)
eovcheck(SleepHrs ~ Deprived, data = sleepDep.df)
onewayPlot(SleepHrs ~ Deprived, data = sleepDep.df)
```

```{r}
normcheck(lm(SleepHrs ~ Deprived, data = sleepDep.df))
cooks20x(lm(SleepHrs ~ Deprived, data = sleepDep.df))
```

```{r}
t.test(SleepHrs ~ Deprived, data = sleepDep.df, var.equal = TRUE)
```

### Methods and Assumption Checks
We are attempting to find the mean difference between two groups of measurements, and as such we need to take a two sample t-test. We assume the measurements are independant.

We checked Equality of Variance, and found that the EOV check for Yes and No appears to show equality of variance, however due to the way measurements were collected (as integer levels) this means that there could be no equality of variance as clusters of values will collapse into a single point. 
The one-way plot we did alongside the EOV check shows that indeed there are clusters of values at certain levels, but equality of variance is roughly maintained, the plots appear to be spread similarly.

We then checked the normality of the sample, and while the Q-Q plot looked slightly abnormal due to the integer levels of measurements, the histogram showed that our assumption of normality is satisfied.

### Executive Summary
We estimate that students who self reported no sleep deprivation slept for an average of 0.51 to 1.48 more hours than those who self reported being sleep deprived.

We estimate that the mean number of hours slept by students who did not self-report feeling sleep deprived is between 0.5 and 1.5 hours more than the mean number of hours slept by students who did. <- denise's explanation